---
title: "Diet and healthcare cost/utilization"
output: github_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

# Required pacakges
pacs <- c(
  "tidyverse", 
  "readxl", 
  "rurality", 
  "tableone",
  "gtsummary",
  "ineq",
  "scales",
  "gt",
  "glmmTMB", 
  "splines", 
  "broom.mixed", 
  "marginaleffects"
)
sapply(pacs, require, character.only = TRUE)

```

## Aim

* To compare healthcare expenditure among five dietary groups of AHS-2 and Medicare fee‑for‑service enrollees

## Datasets

```{r data_prep}

# Medicare crosswalk ------------------------------------------------------

# Crosswalk file: n = 70,968
crosswalk <- read_fwf(
  "./Data/12172/2022/ssn_bene_xwalk_res000058038_req012172_2022.dat",
  fwf_widths(
    c(9, 15, 1, 1, 1),
    c("ORIG_SSN", "BENE_ID", "SSN_MATCH", "SEX_MATCH", "DOB_MATCH")
  )
)

# SSN matched: n = 52,704
all_matched_bene_ids <- crosswalk %>%
  filter(SSN_MATCH == 1)

# Gender or DOB mismatch: n = 3,368
mismatches <- crosswalk %>%
  filter(SSN_MATCH == 1 & (SEX_MATCH == 0 | DOB_MATCH == 0)) %>%
  select(BENE_ID)

# 2 duplicate beneficiary IDs
dup_BENE_IDs <- all_matched_bene_ids %>%
  group_by(BENE_ID) %>%
  summarize(n = sum(n())) %>%
  filter(n > 1)

# Duplicate SSN: n = 106
dup_SSNs <- all_matched_bene_ids %>%
  group_by(ORIG_SSN) %>%
  summarize(n = sum(n())) %>%
  filter(n > 1)

# These will be excluded
exclude_BENE_IDs <- dup_BENE_IDs %>%
  select(BENE_ID) %>%
  union(all_matched_bene_ids %>%
          filter(ORIG_SSN %in% dup_SSNs$ORIG_SSN) %>%
          select(BENE_ID)
  ) %>%
  union(mismatches) 


# Read MBSF data ----------------------------------------------------------

# Read data format for MBSF
fts_mbsf <- read_excel("./data/mbsf_format.xlsx")

# Read data from 2008 to 2020
year <- 2008:2020
fname <- paste0("./Data/12172/", year, "/mbsf_abcd_summary_res000058038_req012172_", year, ".dat")

# Read MBSF files 2008-2020, excluding above beneficiary IDs
all_mbsf <- fname %>%
  lapply(\(x) read_fwf(x, fwf_widths(fts_mbsf$length, fts_mbsf$long_name))) %>%
  lapply(\(x) anti_join(x, exclude_BENE_IDs)) %>%
  setNames(year)

# Read data from 2021 to 2022
year <- 2021:2022
fname <- paste0("./Data/14345/", year, "/mbsf_abcd_summary_res000058038_req014345_", year, ".dat")

add_mbsf <- fname %>%
  lapply(\(x) read_fwf(x, fwf_widths(fts_mbsf$length, fts_mbsf$long_name))) %>%
  lapply(\(x) anti_join(x, exclude_BENE_IDs)) %>%
  setNames(year)

# Long-format over 15 years: n = 463,648
# 46,897 distinct beneficiary IDs
all_mbsf_long <- c(all_mbsf, add_mbsf) %>%
  do.call(rbind, .) %>%
  arrange(BENE_ID, BENE_ENROLLMT_REF_YR)


# Read chronic condition data ---------------------------------------------

# Read data format for CC 2008-2020
fts_cc <- read_excel("./Data/mbsf_cc_format.xlsx")

year <- 2008:2020
fname <- paste0("./Data/12172/", year, "/mbsf_cc_summary_res000058038_req012172_", year, ".dat")

all_cc <- fname %>%
  lapply(\(x) read_fwf(x, fwf_widths(fts_cc$length, fts_cc$long_name))) %>%
  lapply(\(x) anti_join(x, exclude_BENE_IDs)) %>%
  setNames(year)

# Read data format for CC 2021-2022
fts_chronic <- read_excel("./Data/mbsf_chronic_format.xlsx")

year <- 2021:2022
fname <- paste0("./Data/14345/", year, "/mbsf_chronic_summary_res000058038_req014345_", year, ".dat")

add_cc <- fname %>%
  lapply(\(x) read_fwf(x, fwf_widths(fts_chronic$length, fts_chronic$long_name))) %>%
  lapply(\(x) anti_join(x, exclude_BENE_IDs)) %>%
  lapply(\(x) rename(x, HYPERL_EVER = HLP_EVER, CHF_EVER = HF_EVER, HYPERT_EVER = HTN_EVER, HYPOTH_EVER = HYPTHYRD_EVER)) %>%
  setNames(year)

# Long-format over 15 years: n = 463,648
# 46,897 distinct beneficiary IDs
all_cc_long <- c(all_cc, add_cc) %>%
  do.call(bind_rows, .) %>%
  arrange(BENE_ID, BENE_ENROLLMT_REF_YR)


# Read MBSF cost use data -------------------------------------------------

# Read all cost use data
# Long-format over 15 years: n = 463,648
# 46,897 distinct beneficiary IDs
source("read_mbsf_costuse.R")
all_cu_long <- read_mbsf_costuse_years(2008:2022) %>% 
  anti_join(exclude_BENE_IDs, by = "BENE_ID") 


# Merge Medicare data -----------------------------------------------------

# n = 463,648 from 46,897 distinct BENE_IDs
all_mdcr_long <- all_mbsf_long %>% 
  inner_join(all_cc_long, by = c("BENE_ID", "BENE_ENROLLMT_REF_YR")) %>% 
  inner_join(all_cu_long, by = c("BENE_ID", "BENE_ENROLLMT_REF_YR")) 


# AHS-2 Medicare link file ------------------------------------------------

# Read link file
# n = 51,917
ahs <- read_csv("./data/MedicareMatches2022.csv")

# Duplicate analysis ID: n = 104
exclude_analysisIDs <- ahs %>%
  count(AnalysisID) %>%
  filter(n > 1) %>%
  select(AnalysisID)

# Remove them: n = 51,485
ahs_dup_removed <- ahs %>% 
  anti_join(exclude_analysisIDs) %>% 
  mutate(AnalysisID = parse_number(AnalysisID)) %>% 
  setNames(tolower(names(ahs)))

# Read AHS-2 baseline data ------------------------------------------------

# Read AHS-2 BQ, not imputed
# n = 96,114
ahsdata <- read.csv("./data/BaselineDataForMedicare20190531.csv", header = TRUE) %>% 
  as_tibble()
names(ahsdata) <- tolower(names(ahsdata))


# Merge AHS-2 and Medicare data -------------------------------------------

# Merge AHS link file and BQ data
# n = 51,485
ahs_link_bq <- ahsdata %>%
  inner_join(ahs_dup_removed, by = "analysisid") 

# Merge AHS and Medicare data: n = 461,615
# From 46,687 distinct Analysis IDs
mdcr_ahs <- all_mdcr_long %>% 
  janitor::clean_names() %>% 
  inner_join(ahs_link_bq, by = "bene_id")

# Opt-outs: n = 395
optout <- read_csv("./data/OptOutAnalysisIDs.csv") %>% setNames("analysisid")

# n = 162 to be excluded
# mdcr_ahs %>% 
#   distinct(analysisid) %>% 
#   semi_join(optout) %>% 
#   nrow()

# Those who live outside US: n = 163
outside_us <- read_csv("./data/outside_us.csv")

# n = 50 to be excluded
# mdcr_ahs %>% 
#   distinct(analysisid) %>% 
#   semi_join(outside_us, by = "analysisid") %>% 
#   nrow()

# Remove opt-outs and those who live outside
# yielding nobs = 459,660 from 46,475 distinct analysisid
mdcr_ahs <- mdcr_ahs %>% 
  anti_join(optout, by = "analysisid") %>% 
  anti_join(outside_us, by = "analysisid")


# Define model variables --------------------------------------------------

# Dietary patterns, vegstat 1-5 are:
vegstat_labels <- c("vegan", "lacto-ovo", "semi", "pesco", "nonveg")

census_regions <- read_csv("./data/census_regions.csv", col_types = "cccc")

# For urban/rural variable...
# Helper function to get the most frequent FIPS code over 12 months
get_mode <- function(x) {
  x <- x[!is.na(x) & x != "99999"]
  if (length(x) == 0) return(NA_character_)
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

# Variable names for state/country FIPS code, Jan to Dec
fips_cols <- paste0("state_cnty_fips_cd_", sprintf("%02d", 1:12))

# Get the mode of the FIPS code over 12 months
# Will be used to identify urban/rural in the code below
fips_mat            <- as.matrix(mdcr_ahs[fips_cols])
mdcr_ahs$fips_modal <- apply(fips_mat, 1, get_mode)

# n.obs = 459,660 from 46,475 distinct Analysis IDs
mdcr_ahs2 <- mdcr_ahs %>% 
  mutate(
    sex        = factor(sex_ident_cd, labels = c("Male", "Female")),
    sex        = relevel(sex, ref = "Female"),
    age        = age_at_end_ref_yr,
    cage       = age - 65,
    birth_yr   = year(ymd(bene_birth_dt)),
    bth_cohort = cut(birth_yr,
                     breaks = c(-Inf, 1919, 1929, 1939, 1949, Inf),
                     labels = c("<1920", "1920s", "1930s", "1940s", "≥1950")),
    rti_race3  = recode(rti_race_cd + 1, 3, 1, 2, 3, 3, 3, 3),
    rti_race3  = factor(rti_race3, labels = c("NHWhite", "Black", "Other")),
    educyou    = parse_number(educyou, na = c("", "*")),
    educat3    = cut(educyou, breaks = c(0, 3, 6, 9)),
    educat3    = factor(educat3, labels = c("HS or less", "Some college", "Col grad")),
    educat3m   = relevel(educat3, ref = "Col grad"),
    marital    = parse_number(marital, na = c("", "*")),
    marital    = recode(marital, "Never", "Married", "Married", "Married", "Div/Wid", "Div/Wid", "Div/Wid"),
    marital    = factor(marital, levels = c("Married", "Never", "Div/Wid")),
    smoke      = parse_number(smoke, na = c("", " ", "*")),
    smkcat     = case_when(
      smokenow == 1 ~ 3,
      smoke > 1 & is.na(smokenow) ~ 2,
      smoke == 1 ~ 1
    ),
    smkcat     = factor(smkcat, labels = c("Never", "Past", "Current")),
    # smkever    = factor(smkcat, labels = c("Never", "Ever", "Ever")),
    alccat     = case_when(
      alcnow == 1 ~ 3,
      alcohol == 2 & (is.na(alcnow) | alcnow == 0) ~ 2,
      alcohol == 1 ~ 1
    ),
    alccat     = factor(alccat, labels = c("Never", "Past", "Current")),
    # alcever    = factor(alcohol, labels = c("Never", "Ever")),
    vegstat    = factor(vegstat, labels = vegstat_labels),
    vegstat    = fct_relevel(vegstat, "pesco", after = 2),      # Change level order, pesco, semi, nonveg
    vegstat2   = relevel(vegstat, ref = "nonveg"),
    cyear      = extract_year - 2008,
    
    # Number of months of both parts A and B coverage
    ab_entitled_months = pmin(bene_hi_cvrage_tot_mons, bene_smi_cvrage_tot_mons),
    
    # Original reason for entitlement
    entlmt_rsn = factor(entlmt_rsn_orig, labels = c("Age", "DIB/ESRD", "DIB/ESRD", "DIB/ESRD")),
    
    # "Died this year" indicator variable
    bene_death_dt  = lubridate::ymd(bene_death_dt),
    died_this_year = !is.na(bene_death_dt) & lubridate::year(bene_death_dt) == extract_year,
    died_this_year = factor(died_this_year, levels = c(FALSE, TRUE), labels = c("No", "Yes")),
    
    # State FIPS
    state_fips = substr(fips_modal, 1, 2),
    
    # Rural-urban code and define urban/rural
    rucc = get_rucc(fips_modal),
    urban_rural = case_when(
      rucc %in% 1:3 ~ "Urban",
      rucc %in% 4:9 ~ "Rural",
      TRUE           ~ NA_character_
    ),
    urban_rural = factor(urban_rural, levels = c("Urban", "Rural"))
  ) %>%
  left_join(census_regions, by = "state_fips") %>%    # Get census regions "region"
  mutate(
    region = factor(region),
    region = relevel(region, ref = "West")
  )
  

# Calculate total payment subject-year, nominal & real --------------------

# Setup
# Identify payment variables: Payment by Medicare
mdcr_payment_vars <- mdcr_ahs2 %>%
  select(matches("_mdcr_pmt$")) %>%
  names()

# Identify payment variables: Payment by beneficiary
bene_payment_vars <- mdcr_ahs2 %>%
  select(matches("_bene_pmt$")) %>%
  names()

# Identify payment variables: Payment by primary insurer
prmry_payment_vars <- mdcr_ahs2 %>%
  select(matches("_prmry_pmt$")) %>%
  names()

# Identify payment variables: Payment by medicare/beneficiary/primary, as well as per diem
payment_vars <- mdcr_ahs2 %>%
  select(matches("_(mdcr|bene|prmry|perdiem)_pmt$")) %>%
  names()

# CPI-U Medical Care data, monthly, seasonally-adjusted
# https://fred.stlouisfed.org/series/CPIMEDSL
cpi_medical_month <- read_csv("./data/CPIMEDSL.csv")
cpi_medical_year <- cpi_medical_month %>%
  mutate(extract_year = year(observation_date)) %>%
  group_by(extract_year) %>%
  summarize(cpi_medical = mean(CPIMEDSL, na.rm = TRUE))

# Calculate deflator to converts every year's dollars into 2022 dollars
ref_year <- 2022
ref_cpi  <- cpi_medical_year$cpi_medical[cpi_medical_year$extract_year == ref_year]

cpi_medical_year <- cpi_medical_year %>%
  mutate(deflator_2022usd = ref_cpi / cpi_medical)

# Calculate total payment subject-year, both nominal and real 
mdcr_ahs3 <- mdcr_ahs2 %>%
  left_join(cpi_medical_year, by = "extract_year") %>%
  mutate(
    total_pmt           = rowSums(across(all_of(payment_vars)), na.rm = TRUE),
    total_pmt_2022usd   = total_pmt * deflator_2022usd,
    total_pmt_2022usd_k = total_pmt_2022usd / 1000
  )


# Inclusion-exclusion criteria --------------------------------------------

# Exclude those years before reaching age 65
# n.obs = 436,425 from 45,076 distinct Analysis IDs
mdcr_ahs_65 <- mdcr_ahs3 %>% 
  filter(age_at_end_ref_yr >= 65)

# Exclude those years of any months of Medicare Advantage enrollment
# n.obs = 281,197 from 35,864 distinct Analysis IDs
mdcr_ahs_65_ffs <- mdcr_ahs_65 %>% 
  filter(bene_hmo_cvrage_tot_mons == 0)

# Exclude those years of residing in US territories: n = 107
# n.obs = 281,146 from 35,859 distinct Analysis IDs
territory_fips <- c("72", "66", "78", "69", "60")  # PR, Guam, USVI, N. Mariana, American Samoa

mdcr_ahs_65_ffs <- mdcr_ahs_65_ffs %>%
  filter(!(state_fips %in% territory_fips))

# Complete dataset
# Variables needed 
var_needed <- c(
  "analysisid",
  "bene_id",
  "vegstat",
  "vegstat2",
  "extract_year",
  "cyear",
  "age",
  "cage",
  "birth_yr",
  "bth_cohort",
  "sex",
  "rti_race3",
  "region",
  "urban_rural",
  "educat3",
  "educat3m",
  "marital",
  "smkcat",
  "alccat",
  "dual_elgbl_mons",
  "ptd_plan_cvrg_mons",
  "entlmt_rsn",
  "died_this_year",
  "ab_entitled_months",
  "total_pmt",
  "total_pmt_2022usd",
  "total_pmt_2022usd_k"
)

# n.obs = 268,301 from 34,254 distinct Analysis IDs
mdcr_ahs_65_ffs <- mdcr_ahs_65_ffs %>% 
  select(all_of(var_needed)) %>% 
  drop_na()

```

* Medicare data, 2008-2022
  * For details on the Medicare data, see the [AHS-2 Medicare Linkage](https://github.com/keijioda/ahs_medicare_linkage/blob/main/summary.md) repository
  
  * Master Beneficiary Summary File (MBSF), 2008-2022
    * See [Data documentation](https://resdac.org/cms-data/files/mbsf-base/data-documentation)
    * Contains beneficiary characteristics and enrollment/eligibility information
    
  * MBSF Cost and Use (CU), 2008-2022
    * See [Data documentation](https://resdac.org/cms-data/files/mbsf-cost-and-use/data-documentation)
      * See also [Baik et al. (2024), Trends in Racial Disparities in Healthcare Expenditures Among Senior Medicare Fee-for-service Enrollees in 2007-2020](https://pubmed.ncbi.nlm.nih.gov/37957537/)
    * Contains payments made by Medicare, the beneficiary, and primary payers for services covered under Parts A, B, and D
    
  * Together, the two files cover 463,648 person-years, representing 46,897 distinct beneficiaries/participants, after excluding:
    * Gender/DOB mismatch with AHS-2 data
    * Duplicate beneficiary IDs and SSNs

  * For a full list of variable used in this analysis, see the [analysis outline](outline.md)

* AHS-2 Baseline questionnaire (not imputed), including n = 96,114 participants

* After merging Medicare and AHS-2 data, there were n = 46,687 participants
  * Participants who opted out from the study (n = 162) and those who live outside the U.S. (n = 50) were further excluded, yielding n = 46,475 participants (459,660 person-years)


## Outcomes

* Total FFS expenditure: Beneficiary + Medicare + Primary (+ Per diem if available) 
* Part A, Part B, and Part D separately?

## Exposure of interest

* Five dietary patterns derived from AHS-2 baseline questionnaire
  * Assumes that dietary patterns remained stable throughout follow-up

## Inclusion-exclusion criteria

* Starting from 459,660 person-years, we excluded:
  * Person-years before the beneficiary turned 65
    * Resulted in 436,425 person-years from 45,076 distinct IDs
  * Person-years with any month of Medicare Advantage enrollment
    * Resulted in 281,197 person-years from 35,864 distinct IDs
  * Person-years residing in a US territory
    * Resulted in 281,146 person-years from 35,859 distinct IDs
  * Those with missing covariate values
    * Resulted in 268,301 person-years from 34,254 distinct IDs
  
## Descriptive tables

### Participant characteristics at baseline

```{r baseline_descriptives, results = 'asis'}

# Baseline characteristics
baseline_data <- mdcr_ahs_65_ffs %>%
  group_by(bene_id) %>%
  slice_min(extract_year, n = 1, with_ties = FALSE) %>%
  ungroup() %>% 
  rename(ffs_entry_age = age) %>% 
  mutate(
    ab_entitled_cat = case_when(
      ab_entitled_months == 0  ~ "0 months",
      ab_entitled_months < 12  ~ "1-11 months",
      ab_entitled_months == 12 ~ "12 months"),
    ptd_cvrg_cat = case_when(
      ptd_plan_cvrg_mons == 0  ~ "0 months",
      ptd_plan_cvrg_mons < 12  ~ "1-11 months",
      ptd_plan_cvrg_mons == 12 ~ "12 months"),
    dual_elgbl_cat = case_when(
      dual_elgbl_mons  == 0  ~ "0 months",
      dual_elgbl_mons  < 12  ~ "1-11 months",
      dual_elgbl_mons  == 12 ~ "12 months"),
 )

# Variables needed for descriptive table 
table_vars <- c(
  "sex",
  "bth_cohort",
  "ffs_entry_age",
  "rti_race3",
  "region",
  "urban_rural",
  "educat3",
  "marital",
  "smkcat",
  "alccat",
  "entlmt_rsn",
  "ab_entitled_cat",
  "ptd_cvrg_cat",
  "dual_elgbl_cat"
)

# baseline_data %>% 
#   CreateTableOne(table_vars, strata = "vegstat", data = ., addOverall = TRUE) %>%
#   print(showAllLevels = TRUE, test = FALSE) %>% 
#   kableone()

baseline_data %>%
  select(all_of(table_vars), vegstat) %>%
  tbl_summary(
    by = vegstat,
    label = list(
      sex           ~ "Sex",
      ffs_entry_age ~ "Age at FFS entry",
      bth_cohort    ~ "Birth cohort",
      rti_race3     ~ "Race/ethnicity",
      region        ~ "Census region",
      urban_rural   ~ "Urban/rural residence",
      educat3       ~ "Education",
      marital       ~ "Marital status",
      smkcat        ~ "Smoking status",
      alccat        ~ "Alcohol use",
      entlmt_rsn    ~ "Medicare entitlement reason",
      ab_entitled_cat ~ "Part A/B entitlement (months)",
      ptd_cvrg_cat    ~ "Part D coverage (months)",
      dual_elgbl_cat  ~ "Dual eligibility (months)"
    ),
    statistic = list(
      ffs_entry_age ~ "{mean} ({sd})" 
    ),
    digits = list(ffs_entry_age ~ c(1, 1))
  ) %>%
  add_overall() %>%
  modify_header(label ~ "**Characteristic**") %>%
  modify_spanning_header(all_stat_cols() ~ "**Vegetarian Status**") %>%
  bold_labels() %>% 
  as_gt() %>%
  gt::as_raw_html(inline_css = TRUE)

```

### Number of beneficiaries by years and diet pattern

```{r N_bene_by_year_diet, results = 'asis'}

# Person-year by vegstat
n_bene_by_year_diet <- mdcr_ahs_65_ffs %>%
  count(extract_year, vegstat) %>%
  group_by(extract_year) %>%
  mutate(pct = sprintf("%d (%.1f%%)", n, 100 * n / sum(n))) %>%
  ungroup() %>%
  select(extract_year, vegstat, pct) %>%
  pivot_wider(names_from = vegstat, values_from = pct) %>%
  left_join(
    mdcr_ahs_65_ffs %>%
      filter(!is.na(vegstat)) %>%
      count(extract_year, name = "Total"),
    by = "extract_year"
  ) %>%
  relocate(Total, .after = extract_year) %>%
  setNames(c("Year", "Total", "Vegan", "Lacto-ovo", "Pesco-veg", "Semi-veg", "Non-veg"))

n_bene_by_year_diet %>%
  gt() %>%
  tab_header(title = "Number of beneficiaries by year and dietary pattern") %>%
  cols_label(Year = "Year", Total = "Total N") %>%
  cols_align(align = "center", columns = -Year) %>%
  tab_footnote(
    footnote = "Values are N (%), with percentages calculated within year.",
    locations = cells_column_labels(columns = c(Vegan, `Lacto-ovo`, `Pesco-veg`, `Semi-veg`, `Non-veg`))
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  ) %>% 
  gt::as_raw_html(inline_css = TRUE)

```

* All diet groups show growth through roughly 2016, then decline thereafter

```{r N_bene_by_year_diet_plot, fig.height = 4, fig.width = 8}
mdcr_ahs_65_ffs %>%
  count(extract_year, vegstat) %>% 
  ggplot(aes(x = extract_year, y = n, color = vegstat)) +
  geom_point() +
  geom_line() +
  scale_color_discrete(
    name = "Dietary group",
    labels = c(
      "vegan"     = "Vegan",
      "lacto-ovo" = "Lacto-ovo vegetarian",
      "pesco"     = "Pesco-vegetarian",
      "semi"      = "Semi-vegetarian",
      "nonveg"    = "Non-vegetarian"
    )
  ) +
  labs(y = "Number of beneficiary-years", x = "Year") +
  theme_bw() +
  theme(legend.position = "bottom")

```

### Summary statistics of total healthcare payment

* Total healthcare payment was adjusted for inflation using the [CPI for medical care](https://fred.stlouisfed.org/series/cpimedsl)
  * Each year's payments were multiplied by a deflator equal to the ratio of the 2022 index value to that year's index value

* Summary statistics of total healthcare payment by dietary group are shown below, with all dollar values expressed in real (2022) dollars
  * Vegans have the highest proportion of zero-payment beneficiary-years (20.3%) and have the lowest mean/median total payment
  * Vegans show the highest Gini coefficient (0.770), indicating the most unequal spending distribution among all diet groups
  * Spending is highly concentrated at the top across all groups
    * The top 5% of beneficiary-years account for 40–44% of total spending, and 
    * The top 1% account for 14–17%
    * Vegans again showing the highest concentration (44.4% and 16.5%, respectively)
  * Despite having the lowest mean spending, vegans show the greatest relative inequality in spending

```{r total_pay_distribution_stats, results = 'asis'}

# Cost-distribution/skewness table by vegstat
# Uses the full beneficiary-year panel (mdcr_ahs_65_ffs)

cost_summary <- mdcr_ahs_65_ffs %>%
  group_by(vegstat) %>%
  summarise(
    n            = n(),
    pct_zero     = 100 * mean(total_pmt_2022usd == 0, na.rm = TRUE),
    mean         = mean(total_pmt_2022usd, na.rm = TRUE),
    sd           = sd(total_pmt_2022usd, na.rm = TRUE),
    median       = median(total_pmt_2022usd, na.rm = TRUE),
    p25          = quantile(total_pmt_2022usd, 0.25, na.rm = TRUE),
    p75          = quantile(total_pmt_2022usd, 0.75, na.rm = TRUE),
    p90          = quantile(total_pmt_2022usd, 0.90, na.rm = TRUE),
    p95          = quantile(total_pmt_2022usd, 0.95, na.rm = TRUE),
    p99          = quantile(total_pmt_2022usd, 0.99, na.rm = TRUE),
    gini         = ineq::Gini(total_pmt_2022usd, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(iqr = p75 - p25)

# Add an "Overall" row (all vegstat groups pooled)
cost_summary_overall <- mdcr_ahs_65_ffs %>%
  summarise(
    vegstat      = "Overall",
    n            = n(),
    pct_zero     = 100 * mean(total_pmt_2022usd == 0, na.rm = TRUE),
    mean         = mean(total_pmt_2022usd, na.rm = TRUE),
    sd           = sd(total_pmt_2022usd, na.rm = TRUE),
    median       = median(total_pmt_2022usd, na.rm = TRUE),
    p25          = quantile(total_pmt_2022usd, 0.25, na.rm = TRUE),
    p75          = quantile(total_pmt_2022usd, 0.75, na.rm = TRUE),
    p90          = quantile(total_pmt_2022usd, 0.90, na.rm = TRUE),
    p95          = quantile(total_pmt_2022usd, 0.95, na.rm = TRUE),
    p99          = quantile(total_pmt_2022usd, 0.99, na.rm = TRUE),
    gini         = ineq::Gini(total_pmt_2022usd, na.rm = TRUE)
  ) %>%
  mutate(iqr = p75 - p25)

cost_summary_full <- bind_rows(cost_summary, cost_summary_overall)

cost_summary_table <- cost_summary_full %>%
  transmute(
    vegstat,
    n,
    `% zero payment`   = sprintf("%.1f", pct_zero),
    `Mean (SD)`        = sprintf("%s (%s)", scales::comma(round(mean)), scales::comma(round(sd))),
    `Median (IQR)`     = sprintf("%s (%s-%s)", scales::comma(round(median)),
                                 scales::comma(round(p25)), scales::comma(round(p75))),
    P90                = scales::comma(round(p90)),
    P95                = scales::comma(round(p95)),
    P99                = scales::comma(round(p99)),
    Gini               = sprintf("%.3f", gini)
  )

# Add columns of tail share of total payment
tail_share <- mdcr_ahs_65_ffs %>%
  group_by(vegstat2) %>%
  summarise(
    p95_cut          = quantile(total_pmt_2022usd, 0.95, na.rm = TRUE),
    p99_cut          = quantile(total_pmt_2022usd, 0.99, na.rm = TRUE),
    total_dollars    = sum(total_pmt_2022usd, na.rm = TRUE),
    dollars_top5pct  = sum(total_pmt_2022usd[total_pmt_2022usd > p95_cut], na.rm = TRUE),
    dollars_top1pct  = sum(total_pmt_2022usd[total_pmt_2022usd > p99_cut], na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    pct_dollars_top5pct = 100 * dollars_top5pct / total_dollars,
    pct_dollars_top1pct = 100 * dollars_top1pct / total_dollars
  ) %>%
  select(vegstat2, pct_dollars_top5pct, pct_dollars_top1pct)

# Overall row, same logic pooling all groups
tail_share_overall <- mdcr_ahs_65_ffs %>%
  summarise(
    vegstat2         = "Overall",
    p95_cut          = quantile(total_pmt_2022usd, 0.95, na.rm = TRUE),
    p99_cut          = quantile(total_pmt_2022usd, 0.99, na.rm = TRUE),
    total_dollars    = sum(total_pmt_2022usd, na.rm = TRUE),
    dollars_top5pct  = sum(total_pmt_2022usd[total_pmt_2022usd > p95_cut], na.rm = TRUE),
    dollars_top1pct  = sum(total_pmt_2022usd[total_pmt_2022usd > p99_cut], na.rm = TRUE)
  ) %>%
  mutate(
    pct_dollars_top5pct = 100 * dollars_top5pct / total_dollars,
    pct_dollars_top1pct = 100 * dollars_top1pct / total_dollars
  ) %>%
  select(vegstat2, pct_dollars_top5pct, pct_dollars_top1pct)

tail_share_full <- bind_rows(tail_share, tail_share_overall)

# Merge into the existing cost distribution table
cost_summary_full <- cost_summary_full %>%
  rename(vegstat2 = vegstat) %>%  
  left_join(tail_share_full, by = "vegstat2")

veg_labels <- c(
  vegan       = "Vegan",
  `lacto-ovo` = "Lacto-ovo vegetarian",
  pesco       = "Pesco-vegetarian",
  semi        = "Semi-vegetarian",
  nonveg      = "Non-vegetarian",
  Overall     = "Overall"
)

veg_order <- c("vegan", "lacto-ovo", "pesco", "semi", "nonveg", "Overall")

cost_summary_pub <- cost_summary_full %>%
  mutate(vegstat2 = factor(vegstat2, levels = veg_order)) %>%
  arrange(vegstat2) %>%
  transmute(
    `Diet group`        = unname(veg_labels[as.character(vegstat2)]),
    N                    = comma(n),
    `% Zero payment`     = sprintf("%.1f", pct_zero),
    `Mean (SD), $`       = sprintf("%s (%s)", comma(round(mean)), comma(round(sd))),
    `Median (IQR), $`    = sprintf("%s (%s-%s)", comma(round(median)), comma(round(p25)), comma(round(p75))),
    `P90, $`             = comma(round(p90)),
    `P95, $`             = comma(round(p95)),
    `P99, $`             = comma(round(p99)),
    `Gini coefficient`   = sprintf("%.3f", gini),
    `% $ from top 5%`    = sprintf("%.1f", pct_dollars_top5pct),
    `% $ from top 1%`    = sprintf("%.1f", pct_dollars_top1pct)
  )

cost_summary_pub %>%
  gt() %>%
  tab_header(
    title = "Total payment distribution by diet group"
  ) %>%
  tab_footnote(
    footnote = "N reflects beneficiary-years, not unique beneficiaries. Gini coefficient includes beneficiary-years with zero payment. Top 5%/1% dollar shares use each group's own percentile cutoffs.",
    locations = cells_column_labels(columns = c(`Gini coefficient`, `% $ from top 5%`, `% $ from top 1%`))
  ) %>%
  cols_align(align = "center", columns = -`Diet group`) %>%
  tab_style(
    style = cell_borders(sides = "top", color = "black", weight = px(1.5)),
    locations = cells_body(rows = `Diet group` == "Overall")
  ) %>%
  tab_options(table.font.size = px(12)) %>% 
  gt::as_raw_html(inline_css = TRUE)

```

* The same statistics were calculated again after excluding beneficiary-years of zero payments. After excluding zero-payment years:
  * Mean and median payments rise substantially for all diet groups
    * The Gini coefficients dropped as well
    * Spending concentration at the top remains high but is somewhat less extreme
  * Vegans retain the highest spending concentration among users (highest Gini, highest top 5%/1% shares)
    * suggesting that even conditional on using any healthcare, vegan spending is somewhat more concentrated among a smaller subset of high-cost users than in other diet groups

```{r total_pay_distribution_stats_exclude_zero, results = 'asis'}

# Cost-distribution/skewness table by vegstat
# Uses the full beneficiary-year panel (mdcr_ahs_65_ffs)

cost_summary <- mdcr_ahs_65_ffs %>%
  filter(total_pmt_2022usd > 0) %>% 
  group_by(vegstat) %>%
  summarise(
    n            = n(),
    pct_zero     = 100 * mean(total_pmt_2022usd == 0, na.rm = TRUE),
    mean         = mean(total_pmt_2022usd, na.rm = TRUE),
    sd           = sd(total_pmt_2022usd, na.rm = TRUE),
    median       = median(total_pmt_2022usd, na.rm = TRUE),
    p25          = quantile(total_pmt_2022usd, 0.25, na.rm = TRUE),
    p75          = quantile(total_pmt_2022usd, 0.75, na.rm = TRUE),
    p90          = quantile(total_pmt_2022usd, 0.90, na.rm = TRUE),
    p95          = quantile(total_pmt_2022usd, 0.95, na.rm = TRUE),
    p99          = quantile(total_pmt_2022usd, 0.99, na.rm = TRUE),
    gini         = ineq::Gini(total_pmt_2022usd, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(iqr = p75 - p25)

# Add an "Overall" row (all vegstat groups pooled)
cost_summary_overall <- mdcr_ahs_65_ffs %>%
  filter(total_pmt_2022usd > 0) %>% 
  summarise(
    vegstat      = "Overall",
    n            = n(),
    pct_zero     = 100 * mean(total_pmt_2022usd == 0, na.rm = TRUE),
    mean         = mean(total_pmt_2022usd, na.rm = TRUE),
    sd           = sd(total_pmt_2022usd, na.rm = TRUE),
    median       = median(total_pmt_2022usd, na.rm = TRUE),
    p25          = quantile(total_pmt_2022usd, 0.25, na.rm = TRUE),
    p75          = quantile(total_pmt_2022usd, 0.75, na.rm = TRUE),
    p90          = quantile(total_pmt_2022usd, 0.90, na.rm = TRUE),
    p95          = quantile(total_pmt_2022usd, 0.95, na.rm = TRUE),
    p99          = quantile(total_pmt_2022usd, 0.99, na.rm = TRUE),
    gini         = ineq::Gini(total_pmt_2022usd, na.rm = TRUE)
  ) %>%
  mutate(iqr = p75 - p25)

cost_summary_full <- bind_rows(cost_summary, cost_summary_overall)

cost_summary_table <- cost_summary_full %>%
  transmute(
    vegstat,
    n,
    `% zero payment`   = sprintf("%.1f", pct_zero),
    `Mean (SD)`        = sprintf("%s (%s)", scales::comma(round(mean)), scales::comma(round(sd))),
    `Median (IQR)`     = sprintf("%s (%s-%s)", scales::comma(round(median)),
                                 scales::comma(round(p25)), scales::comma(round(p75))),
    P90                = scales::comma(round(p90)),
    P95                = scales::comma(round(p95)),
    P99                = scales::comma(round(p99)),
    Gini               = sprintf("%.3f", gini)
  )

# Add columns of tail share of total payment
tail_share <- mdcr_ahs_65_ffs %>%
  filter(total_pmt_2022usd > 0) %>% 
  group_by(vegstat2) %>%
  summarise(
    p95_cut          = quantile(total_pmt_2022usd, 0.95, na.rm = TRUE),
    p99_cut          = quantile(total_pmt_2022usd, 0.99, na.rm = TRUE),
    total_dollars    = sum(total_pmt_2022usd, na.rm = TRUE),
    dollars_top5pct  = sum(total_pmt_2022usd[total_pmt_2022usd > p95_cut], na.rm = TRUE),
    dollars_top1pct  = sum(total_pmt_2022usd[total_pmt_2022usd > p99_cut], na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    pct_dollars_top5pct = 100 * dollars_top5pct / total_dollars,
    pct_dollars_top1pct = 100 * dollars_top1pct / total_dollars
  ) %>%
  select(vegstat2, pct_dollars_top5pct, pct_dollars_top1pct)

# Overall row, same logic pooling all groups
tail_share_overall <- mdcr_ahs_65_ffs %>%
  filter(total_pmt_2022usd > 0) %>% 
  summarise(
    vegstat2         = "Overall",
    p95_cut          = quantile(total_pmt_2022usd, 0.95, na.rm = TRUE),
    p99_cut          = quantile(total_pmt_2022usd, 0.99, na.rm = TRUE),
    total_dollars    = sum(total_pmt_2022usd, na.rm = TRUE),
    dollars_top5pct  = sum(total_pmt_2022usd[total_pmt_2022usd > p95_cut], na.rm = TRUE),
    dollars_top1pct  = sum(total_pmt_2022usd[total_pmt_2022usd > p99_cut], na.rm = TRUE)
  ) %>%
  mutate(
    pct_dollars_top5pct = 100 * dollars_top5pct / total_dollars,
    pct_dollars_top1pct = 100 * dollars_top1pct / total_dollars
  ) %>%
  select(vegstat2, pct_dollars_top5pct, pct_dollars_top1pct)

tail_share_full <- bind_rows(tail_share, tail_share_overall)

# Merge into the existing cost distribution table
cost_summary_full <- cost_summary_full %>%
  rename(vegstat2 = vegstat) %>%  
  left_join(tail_share_full, by = "vegstat2")

cost_summary_pub <- cost_summary_full %>%
  mutate(vegstat2 = factor(vegstat2, levels = veg_order)) %>%
  arrange(vegstat2) %>%
  transmute(
    `Diet group`        = unname(veg_labels[as.character(vegstat2)]),
    N                    = comma(n),
    `% Zero payment`     = sprintf("%.1f", pct_zero),
    `Mean (SD), $`       = sprintf("%s (%s)", comma(round(mean)), comma(round(sd))),
    `Median (IQR), $`    = sprintf("%s (%s-%s)", comma(round(median)), comma(round(p25)), comma(round(p75))),
    `P90, $`             = comma(round(p90)),
    `P95, $`             = comma(round(p95)),
    `P99, $`             = comma(round(p99)),
    `Gini coefficient`   = sprintf("%.3f", gini),
    `% $ from top 5%`    = sprintf("%.1f", pct_dollars_top5pct),
    `% $ from top 1%`    = sprintf("%.1f", pct_dollars_top1pct)
  )

cost_summary_pub %>%
  gt() %>%
  tab_header(
    title = "Total payment distribution by diet group"
  ) %>%
  tab_footnote(
    footnote = "N reflects beneficiary-years, not unique beneficiaries. Gini coefficient excludes beneficiary-years with zero payment. Top 5%/1% dollar shares use each group's own percentile cutoffs.",
    locations = cells_column_labels(columns = c(`Gini coefficient`, `% $ from top 5%`, `% $ from top 1%`))
  ) %>%
  cols_align(align = "center", columns = -`Diet group`) %>%
  tab_style(
    style = cell_borders(sides = "top", color = "black", weight = px(1.5)),
    locations = cells_body(rows = `Diet group` == "Overall")
  ) %>%
  tab_options(table.font.size = px(12)) %>% 
  gt::as_raw_html(inline_css = TRUE)

```

## Exploratory analyses of healthcare utilization and payment

### Distribution of annual total healthcare payment

* The histogram below shows the annual total payment in 2022 real dollars
  * The x-axis uses a pseudo-log scale to display the large mass of zero-payment beneficiary-years (no healthcare usage) alongside the positive-payment distribution

```{r distribution_total_payment, fig.height = 4, fig.width = 5}

# Histogram of annual total payment, real, at person-year level

minor_breaks <- as.vector(outer(1:9, 10^(0:5)))
dollar_breaks <- c(0, 10, 100, 1000, 10000, 100000, 1000000)
dollar_labels <- c("$0", "$10", "$100", "$1,000", "$10K", "$100K", "$1M")

mdcr_ahs_65_ffs %>% 
  ggplot(aes(x = total_pmt_2022usd)) +
  geom_histogram(bins = 40) +
  scale_x_continuous(
    transform = scales::pseudo_log_trans(),
    breaks = dollar_breaks,
    labels = dollar_labels,
    minor_breaks = minor_breaks,
    limits = c(-.5, 2E6)
  ) +
  labs(x = "Annual total payment (2022 $), pseudo-log scale", y = "Frequency") +
  theme_bw()

```


### Crude utililzation rate by dietary pattern

* The plot below shows the crude (unadjusted) proportion of beneficiaries with any healthcare payment, by year and dietary pattern
  * Vegans consistently show the lowest crude utilization rate across the entire study
    * Semi-vegetarians show the highest utilization rate
  * All diet groups show a declining trend in utilization over time -- Why?
    * Models need to adjust for calendar year

```{r healthcare_utilization_proportion, fig.height = 4, fig.width = 8}
# Crude utilization rate by dietary group ---------------------------------

# Percent of zero total payment = 16.4%
# Over years
mdcr_ahs_65_ffs %>%
  mutate(
    zero_pmt = ifelse(total_pmt_2022usd == 0, 0, 1)
  ) %>%
  group_by(extract_year, vegstat) %>%
  summarize(
    pct_use = mean(zero_pmt, na.rm = TRUE)
  ) %>% 
  ggplot(aes(x = extract_year, y = pct_use, color = vegstat)) +
  geom_point() +
  geom_line() + 
  scale_y_continuous(
    labels = scales::percent,
    limits = c(.75, .91)
  ) +
  scale_color_discrete(
    name = "Dietary group",
    labels = c(
      "vegan"     = "Vegan",
      "lacto-ovo" = "Lacto-ovo vegetarian",
      "pesco"     = "Pesco-vegetarian",
      "semi"      = "Semi-vegetarian",
      "nonveg"    = "Non-vegetarian"
    )
  ) +
  labs(
    x = "Year",
    y = "Healthcare utilization (% with any payment)"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")

```


### Medicare Advantage enrollment by dietary pattern

* The plot below shows the proportion of beneficiaries with any MA coverage months by dietary pattern
  * MA enrollment rose sharply across all dietary groups, roughly doubling in most groups
    * Vegans and lacto-ovos consistently had the lowest MA uptake
    * while pesco- & non-vegetarians consistently had the highest
  * The inflection point around 2016–2017 coincides with the steepest decline observed in the crude FFS utilization rate above 
    * suggesting that a meaningful part of that declining trend reflects a shrinking, increasingly selected FFS sample rather than a true drop in healthcare need

```{r MA_enrollment, fig.height = 4, fig.width = 8}

# Medicare advantage enrollment by dietary pattern
mdcr_ahs %>% 
  filter(!is.na(vegstat)) %>% 
  filter(!(state_code %in% territory_fips)) %>% 
  mutate(
    ma = ifelse(bene_hmo_cvrage_tot_mons > 0, 1, 0),
    vegstat    = factor(vegstat, labels = vegstat_labels),
    vegstat    = fct_relevel(vegstat, "pesco", after = 2)
  ) %>% 
  group_by(extract_year, vegstat) %>% 
  summarize(pct_ma = mean(ma)) %>% 
  ggplot(aes(x = extract_year, y = pct_ma, color = vegstat)) +
  geom_point() +
  geom_line() +
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0, 0.6)
  ) +
  scale_color_discrete(
    name = "Dietary group",
    labels = c(
      "vegan"     = "Vegan",
      "lacto-ovo" = "Lacto-ovo vegetarian",
      "pesco"     = "Pesco-vegetarian",
      "semi"      = "Semi-vegetarian",
      "nonveg"    = "Non-vegetarian"
    )
  ) +
  labs(
    x = "Year",
    y = "Proportion of beneficiaries with Medicare Advantage"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")

```


### Unadjusted trends in total healthcare expenditure by dietary pattern

* Mean annual total healthcare expenditure by dietary pattern over time
  * Includes beneficiary-years with zero payment
  * Vegans consistently had the lowest mean annual healthcare expenditure throughout the study period
    * Non-vegetarians and semi-vegetarians consistently had the highest
  * Nominal dollars show a clear upward trend over time
    * The trend flattens substantially in 2022-real dollars
    * suggesting the nominal increase largely reflects inflation rather than real growth in utilization
  * Note the dip in 2020, likely related to the COVID-19 pandemic

```{r total_paymnt_by_year_vegstat, fig.height = 4, fig.width = 9}

# Mean annual total payment
# Compare nominal and 2022 real $
mdcr_ahs_65_ffs %>% 
  group_by(extract_year, vegstat) %>%
  summarize(
    total_pmt = mean(total_pmt),
    total_pmt_2022usd = mean(total_pmt_2022usd)
  ) %>%
  pivot_longer(-(1:2), names_to = "var", values_to = "total_pmt") %>% 
  ggplot(aes(x = extract_year, y = total_pmt, color = vegstat)) +
  geom_point() +
  geom_line() +
  labs(
    x = "Year",
    y = "Mean annual total healthcare expenditure ($)"
  ) +
  scale_y_continuous(limits = c(0, NA)) +
  scale_color_discrete(
    name = "Dietary group",
    labels = c(
      "vegan"     = "Vegan",
      "lacto-ovo" = "Lacto-ovo vegetarian",
      "pesco"     = "Pesco-vegetarian",
      "semi"      = "Semi-vegetarian",
      "nonveg"    = "Non-vegetarian"
    )
  ) +
  facet_wrap(
    ~ var,
    labeller = labeller(var = c(
      total_pmt         = "Nominal dollars",
      total_pmt_2022usd = "2022 real dollars"
    ))
  ) +
  theme_bw() +
  theme(legend.position = "bottom")

```



## Modelling approaches

### Two-part hurdle model

* Let $Y_{ij}$ denote the total medical expenditure for subject $i\,(i = 1, \cdots, n)$ on year $j \,(j = 2008,\cdots,2022)$
  * This requires models that account for correlations among repeated measurements $Y_{ij}$ within the subject $i$ over $j = 2008,\cdots,2022$
* The total medical expenditure $Y_{ij}$ is expected to have many zeros (beneficiaries with no healthcare usage)
  * For the remaining beneficiaries,  $Y_{ij}$ is positive and expected to be highly right-skewed
* We use a two-part hurdle model:
  * The first part models $Pr(Y_{ij} > 0)$, the probability that a beneficiary has any positive healthcare spending
    * Fit using a generalized linear mixed model (GLMM) with a binomial distribution
    * Yields odds ratios for the association between diet and the likelihood of healthcare utilization
  * The second part models $Y_{ij} \mid Y_{ij} > 0$, the amount of spending among beneficiaries with positive spending
    * Fit using a GLMM with a Gamma distribution
    * Yields ratios of means for the association between diet and healthcare cost, conditional on positive spending
  * Expected healthcare spending is obtained by multiplying the predicted probabilities of positive spending, $Pr(Y_{ij} > 0)$, from the first part, by the predicted spending conditional on positive spending $Y_{ij} \mid Y_{ij} > 0$, from the second part

### Covariates
* Covariates:
  * Calendar year, centered at 2008
  * Demographics:
    * Age at the end of year, centered at 65 yo (time-dependent)
    * Gender
    * RTI race
    * Census region
    * Urban/rural (from beneficiary's State/County FIPS code)
    * AHS-2: Education
    * AHS-2: Marital status
  * Lifestyle
    * AHS-2: Smoking status
    * AHS-2: Alcohol use
  * Medicare
    * Dual eligibility, number of months
    * Part D coverage, number of months
    * Original reason for entitlement (Age or disability/ESRD)
    * Indicator for beneficiaries' final year of life
    